---
title: "Comparison between RNA-seq and Immunohistochemistry data"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    code_folding: show
    fig_height: 8
    fig_width: 10
    highlight: haddock
    theme: readable
    toc: yes
    toc_depth: 3
    toc_float: yes
---

In the SHIVA trial, estrogen, androgen and progesterone status (ER, AR and PR) were collected in order to associate the patients to a hormone therapy. The hormone receptors ER, AR and PR were collected using Immunohistochemistry (IHC). Immunohistochemestry values are available through TCGA for a limited number of samples but RNA-Seq could represent a valid proxy. In this section we want to explore the relationship between RNA-seq and Immunohistochemistry and, possibly, identify a threshold that we can use in the simulation to appropriately identify over-expressed samples. IHC categories for ER will be compared to *ESR1* expression values and IHC PR categories to the RNA values of *PGR* gene.

Comparative analysis
================================================================================

To run the comparison analysis we will need two datasets:

* Dataset with IHC values, 
* Dataset with RNA-seq expression values

Both dataset need to have in common the same patients so that we can reconstruct the index.

The analysis will be run on two hormone receptors: 

* ER
* PR


ER analysis 
--------------------------------------------------------------------------------

### IHC dataset

As Input dataset we are choosing to use: 

Clinical data downloaded from cBioportal for the dataset: **Breast Invasive Carcinoma (TCGA, Cell 2015)** - [LINK](https://git.ieo.eu/acc-bioinfo/meta/activity)

The dataset has been downloaded and stored as  ```brca_tcga_pub2015_clinical_data.tsv```. 

```{r, message=FALSE, warning=FALSE}
suppressMessages(library(dplyr))
suppressMessages(library(ggplot2))
suppressMessages(library(plotly))
suppressMessages(library(readr))
suppressMessages(library(knitr))
suppressMessages(library(PrecisionTrialDesigner))

ihc <-readr::read_tsv("../external_resources/brca_tcga_pub2015_clinical_data.tsv")
ihcFilter <- ihc %>% 
  dplyr::select(`Patient ID`
                , `ER Status By IHC`
                , `ER Status IHC Percent Positive`
                ) %>%
  dplyr::filter(!is.na(`ER Status By IHC`)) %>% # Remove the <NA>
  dplyr::filter(!is.na(`ER Status IHC Percent Positive`)) %>% # Remove the <NA>
  dplyr::rename(case_id=`Patient ID`, er_status=`ER Status By IHC`, ihc_value=`ER Status IHC Percent Positive`)
 
# preview
kable(head(ihcFilter), caption="top 6 rows")
```


### RNA-seq dataset

The RNA-seq dataset was extracted using PTD function.!

```{r}
panel_design <- data.frame(drug=""
    , gene_symbol="ESR1"
    , alteration="expression"
    , exact_alteration="up"
    ,	mutation_specification=""
    ,	group="")


panel <- newCancerPanel(panel_design)
panel <- getAlterations(panel, tumor_type = "brca_tcga")
panel <- subsetAlterations(panel)

# Load data from SHIVA retrospective analaysis
#panel <- readRDS("../Temp/shiva_panel.rds")

# Fetch data
rnaseq <- panel@dataFull$expression$data %>%
  filter(tumor_type == "brca") %>%
  filter(gene_symbol == "ESR1") %>%
  select(case_id, expressionValue)

# Preview
kable(head(rnaseq), caption = "top 6 rows")
```


### Inner Join datasets

Preview the results from the inner join.

```{r}
df <- dplyr::inner_join(rnaseq, ihcFilter, by="case_id")
# preview
kable(head(df
          # ADD BUTTONS TO THE TABLE
          , extensions = 'Buttons'
          , options = list(
               dom = 'lBfrtip'
              , buttons = c('copy', 'csv', 'excel')
              )
          , caption = "Comparison between Missing and Submitted regions (bp) in the panel"
          ))
```


### Comparison Analysis



#### Explore RNA-seq Z-score 

```{r}
# explore z-score value
p1 <- ggplot(df, aes(x=expressionValue)) +
        geom_density(kernel="gaussian") + 
        geom_vline(aes(xintercept=0.3, color="red")) + 
        labs(x="Expression z-scores", title="Rna-seq expression density plot") + 
        theme(legend.position = "none", plot.title=element_text(size=10))
p1


ggsave(filename="../Figures/fig_extra1.svg", plot=p1, device = "svg")
```

#### Explore ICH values

```{r}
# barplot 
p2 <- ggplot(data=df, aes(x=ihc_value)) + 
        geom_bar(stat = "count", position = "stack") + 
        labs(title="Barplot with COUNT of patients in each ER ICH expression value (from 0 to 100%)")+
        theme(legend.position = "none", plot.title=element_text(size=10))
p2

ggsave(filename="../Figures/fig_extra2.svg", plot=p2, device = "svg")
```


#### Compare

```{r, fig.width=7}
p3 <- ggplot(data=df, aes(x=ihc_value, y=expressionValue, group=1)) +
    geom_point(colour="red", size=1, shape=21, fill="white") +
    labs(title="Comparison between RNA-seq and IHC values for ER in Breast cancer") +
    xlab("IHC value") +
    ylab("RNA-seq z-score") +
    geom_smooth(method="lm") +
    geom_hline(yintercept =0.3) +
    theme(legend.position = "none", plot.title=element_text(size=10))

#ggplotly(p3,width = 650, height = 400, margin(t=1000))

ggsave(filename="../Figures/fig_extra3.svg", plot=p3, device = "svg")
```


#### Fit to a linear model


```{r}
# Convert chategorical values to continue numerical value 
# <10% = 1
# 10-19% = 2
# etc..
df$ihc_value2 <- as.numeric(factor(df$ihc_value))
# Fit the data into a linea regressino model ache chek the coefficients
summary(lm(df$expressionValue ~ ihc_value2, df))
```

There is a significant linear relationship between the predictor and the outcome. Although the $R^2$ value is very low ($R^2$ indicates the percentage of total variation explained by the linear relationship with the predictors). 

*  Pearson correlation: **`r cor(as.numeric(df$ihc_value2), df$expressionValue)`**

### Put all the plots together


```{r, fig.height=8, fig.width=12, echo=FALSE}
library(grid)
library(gridExtra)
grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 2)))

print(p1 + coord_flip(), vp = viewport(layout.pos.row = 1, layout.pos.col = 1))
print(p3, vp = viewport(layout.pos.row = 1, layout.pos.col = 2))
print(p2, vp = viewport(layout.pos.row = 2, layout.pos.col = 2))
```


PR analysis
--------------------------------------------------------------------------------

### IHC dataset

```{r}
ihcPRFilter <- ihc %>% 
  dplyr::select(`Patient ID`
                , `PR status by ihc`
                , `PR status ihc percent positive`
                ) %>%
  dplyr::filter(!is.na(`PR status by ihc`)) %>% # Remove the <NA>
  dplyr::filter(!is.na(`PR status ihc percent positive`)) %>% # Remove the <NA>
  dplyr::rename(case_id=`Patient ID`, pr_status=`PR status by ihc`, ihc_value=`PR status ihc percent positive`)
```

### RNA-seq dataset

The RNA-seq dataset was extracted using PTD function.!

```{r}
panel_design <- data.frame(drug=""
    , gene_symbol="PGR"
    , alteration="expression"
    , exact_alteration="up"
    ,	mutation_specification=""
    ,	group="")


panel <- newCancerPanel(panel_design)
panel <- getAlterations(panel, tumor_type = "brca_tcga")
panel <- subsetAlterations(panel)

# Load data from SHIVA retrospective analaysis
# Fetch data
rnaseq_PR <- panel@dataFull$expression$data %>%
  filter(tumor_type == "brca") %>%
  filter(gene_symbol == "PGR") %>%
  select(case_id, expressionValue)
```

### Inner join datasets

```{r}
# join
dfPR <- dplyr::inner_join(rnaseq_PR, ihcPRFilter, by="case_id")
```


### Comparison analysis

#### Explore RNA-seq Z-score 

```{r}
p1 <- ggplot(dfPR, aes(x=expressionValue)) +
        geom_density(kernel="gaussian") + 
        geom_vline(aes(xintercept=0.3, color="red")) + 
        labs(x="Expression z-scores", title="Rna-seq expression density plot") + 
        theme(legend.position = "none", plot.title=element_text(size=10))
p1
```


#### Explore ICH values

```{r}
p2 <- ggplot(data=dfPR, aes(x=ihc_value)) + 
        geom_bar(stat = "count", position = "stack") + 
        labs(title="Barplot with COUNT of patients in each PR ICH expression value (from 0 to 100%)")+
        theme(legend.position = "none", plot.title=element_text(size=10))
p2
```



#### Compare

```{r, fig.width=7}
p3 <- ggplot(data=dfPR, aes(x=ihc_value, y=expressionValue, group=1)) +
    geom_point(colour="red", size=1, shape=21, fill="white") +
    labs(title="Comparison between RNA-seq and IHC values for PR in Breast cancer") +
    xlab("IHC value") +
    ylab("RNA-seq z-score") +
    geom_smooth(method="lm") +
    geom_hline(yintercept =0.3) +
    theme(legend.position = "none", plot.title=element_text(size=10))

#ggplotly(p3,width = 650, height = 400, margin(t=1000))
p3
```


#### Fit to a linear model

```{r}
# Convert chategorical values to continue numerical value 
# <10% = 1
# 10-19% = 2
# etc..
dfPR$ihc_value2 <- as.numeric(factor(dfPR$ihc_value))
# Fit the data into a linea regressino model ache chek the coefficients
summary(lm(expressionValue ~ ihc_value2, dfPR))
```

### Put all the plots together


```{r, fig.height=8, fig.width=12, echo=FALSE}
library(grid)
grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 2)))

print(p1 + coord_flip(), vp = viewport(layout.pos.row = 1, layout.pos.col = 1))
print(p3, vp = viewport(layout.pos.row = 1, layout.pos.col = 2))
print(p2, vp = viewport(layout.pos.row = 2, layout.pos.col = 2))

```

# Session Info

```{r}
sessionInfo()
```

